Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
typos

This procedure is iterated till a fixed point is reached. In this
example this is reached in the first step. The final spans
are used to construct stencil specification, where
each span is product \texttt{*} of regions and multiple spans are
combined with the region sum \texttt{+}.

@@

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \spanOp which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional space. That is, we compute
 $(\mu \; \spanOp) N$ where \spanOp
is defined as follows for each variable in the map:
%
\begin{enumerate}

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be un-permuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

@@

Next, the set of spans $\spanOp(M)$ covering the indexing
space is converted into abstract syntax trees 
of the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}), and which is wrapped
with the $\textsf{Approx}$ type (see
Definition~\ref{def:mult-and-approx},
p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
for brevity.  The algebra mirrors the shape
of the region combinators for addition and multiplication, 
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
  \mathsf{A}(\specDNF) \\
\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
        \mathsf{A}(\specDNF) \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification 
which contains a one-dimensional region. 
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by by $\oplus$, that is, this step is given by:
\begin{equation*}
\textsf{spec}(M) = 
\sum\limits_{\vect{x} \in \spanOp(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).
We unpack the definitions of this algebra.

\begin{definition}[Specifications in DNF]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
The set \specDNF{} is a subset of syntax trees \textit{spec}
of specifications in disjunctive normal form over regions 
where the disjunct is \term{+} (since its model is in terms of set
union) and the conjunct is \term{*} (since its model is a form of tensor).
That is elements of \specDNF{} are of the form:
%%
\[
(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
\]
%%
More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
%%
\begin{align*}
\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
  \tyProd{\specDNF} \\
\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
   \textit{region}
\end{align*}
%%
We incldue the empty element $\epsilon$ which is not present in the
syntax. 
In the following, $s, t$ range over elements of
\specDNF{}, $\tySum{s}, \tySum{t}$ range over elements of 
$\tySum{\specDNF}$ and $\tyProd{s}, \tyProd{t}$ range over elements
of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
\ldots \term{*} \; r^1_n$).
\end{definition}

@@

A specification is first synthesised into its model. Then occurrences of an
array in an assignment are checked against the model for consistency as defined
in Figure~\ref{fig:consistency}. The consistency predicate is defined
recursively.

@@

Next, consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every indexing scheme in the
model generalises some relativised access pattern that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns can be generalised with
the indexing schemes of the model. Hence, the actual cells are a subset of the
specified ones. In combination an exact specification, \ie{} one without any
approximation modifiers, is one that is both a lower and an upper bound.

@@

We will consider an extended example of a stencil computation that is
consistent with its annotation to show how checking works. The following is an
example with a looser but still consistent version of the \textit{fivepoint}
stencil specification inferred previously in the
Subsection~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
\begin{enumerate}
\item Generate model from the specification as described in
  Section~\ref{sec:semantics}.
%
\begin{dmath*}
\interp{\texttt{stencil readOnce, fivepoint}}_2 =
  \textsf{once}(
    \textsf{exact}(
      \{ \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
          \vtwo{\infty}{1}\}))
\end{dmath*}
%
\item Now $cons$ function is applied with this model along side with the
  assignment on line three and array $b$. For brevity the assignment is
  referred as $L3$ and the set of unifiers defined in the previous step as
  $m$.

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
%
\item $cons$ predicate from the previous step evaluates to
  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
%
  The former condition holds as no access pattern is repeated in $L3$ by simple
  inspection.
%
\item We further expand the function to 
  \begin{equation*}
    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
    cons(\textsf{mult}(\textsf{low}(m),L3,b).
  \end{equation*}
%
\item To evaluate the two predicates above, we need indexing schemes for the
  array in question. This is obtained by $analyse$ function from
  Subsection~\ref{subsec:inference}.

@@

Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The first four steps will proceed the same way. In the fifth step, result
of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In the
sixth step, since the model has not changed and the indexing schemes are only
augmented with a new element, the generalisations for lower bound still hold.
However, there are no elements in the model, $m$, that can generalise
$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification was as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
been needed. This semantically makes sense since the region
$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.

>>>>

This procedure is iterated until a fixed point is reached. In this
example this is reached in the first step. The final spans
are used to construct stencil specification, where
each span is product \texttt{*} of regions and multiple spans are
combined with the region sum \texttt{+}.

@@

The algorithm to create contiguous spans, covering the space of
indices, now proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then computed for the following
procedure \spanOp which, per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional space. That is, we compute
 $(\mu \; \spanOp) N$ where \spanOp
is defined as follows for each variable in the map:
%
\begin{enumerate}

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$) we pair the
permutation function with the set of permuted spans so that
the spans can be un-permuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

@@

Next, the set of spans $\spanOp(M)$ covering the indexing
space is converted into abstract syntax trees 
of the specification language. This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}), and which is wrapped
with the $\textsf{Approx}$ type (see
Definition~\ref{def:mult-and-approx},
p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
for brevity.  The algebra mirrors the shape
of the region combinators for addition and multiplication, 
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
  \mathsf{A}(\specDNF) \\
\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
        \mathsf{A}(\specDNF) \\
\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
\end{align*}
%%
The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
values (one-dimensional) and constructs a specification 
which contains a one-dimensional region. 
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) = 
\sum\limits_{\vect{x} \in \spanOp(M)}
\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
(where $\oplus$ is the additive and $\odot$ is the multiplicative used
within $\Sigma$ and $\Pi$ respectively here).
We unpack the definitions of this algebra.

\begin{definition}[Specifications in DNF]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
The set \specDNF{} is a subset of syntax trees \textit{spec}
of specifications in disjunctive normal form over regions 
where the disjunct is \term{+} (since its model is in terms of set
union) and the conjunct is \term{*} (since its model is a form of tensor).
That is elements of \specDNF{} are of the form:
%%
\[
(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
\]
%%
More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
%%
\begin{align*}
\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
  \tyProd{\specDNF} \\
\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
   \textit{region}
\end{align*}
%%
We include the empty element $\epsilon$ which is not present in the
syntax. 
In the following, $s, t$ range over elements of
\specDNF{}, $\tySum{s}, \tySum{t}$ range over elements of 
$\tySum{\specDNF}$ and $\tyProd{s}, \tyProd{t}$ range over elements
of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
\ldots \term{*} \; r^1_n$).
\end{definition}

@@

A specification is first synthesised into its model. Then the occurrences of an
array in an assignment are checked against the model for consistency as defined
in Figure~\ref{fig:consistency}. The consistency predicate is defined
recursively.

@@

Next, a consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every indexing scheme in the
model generalises some relativised access pattern that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns can be generalised with
the indexing schemes of the model. Hence, the actual cells are a subset of the
specified ones. In combination an exact specification, \ie{} one without any
approximation modifiers, is one that is both a lower and an upper bound.

@@

We will consider an extended example of a stencil computation that is
consistent with its annotation to show how checking works. The following is an
example with a looser but still consistent version of the \textit{fivepoint}
stencil specification inferred previously in the
Subsection~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
\begin{enumerate}
\item Generate model from the specification as described in
  Section~\ref{sec:semantics}.
%
\begin{dmath*}
\interp{\texttt{stencil readOnce, fivepoint}}_2 =
  \textsf{once}(
    \textsf{exact}(
      \{ \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
          \vtwo{\infty}{1}\}))
\end{dmath*}
%
\item Now the $cons$ predicate is applied with this model alongside the
  assignment on line three and array $b$. For brevity the assignment is
  referred to as $L3$ and the set of unifiers defined in the previous step as
  $m$.

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
%
\item $cons$ predicate from the previous step evaluates to
  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
%
  This condition holds as no access pattern is repeated in $L3$ by simple
  inspection.
%
\item We further expand the function to 
  \begin{equation*}
    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
    cons(\textsf{mult}(\textsf{low}(m),L3,b).
  \end{equation*}
%
\item To evaluate the two predicates above, we need indexing schemes for the
  array in question. This is obtained by the $analyse$ function from
  Subsection~\ref{subsec:inference}.

@@

Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The first four steps proceed as before. In the fifth step, the result
of $\mathit{analyse}(L3',b)$ will have an extra term $\vtwoh{2}{2}$. In the
sixth step, since the model has not changed and the indexing schemes are only
augmented with a new element, the generalisations for lower bound still hold.
However, there are no elements in the model, $m$, that can generalise
$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
been needed. This semantically makes sense since the region
$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.
